On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-23 tex/draft/draft2.tex
draft4 tex + pdf, but several other files added automatically when trying to merge the pdf

Standard calculations that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume a steady state at all times and find that, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass (see section \ref{intro}). In our model the atmosphere is no longer in steady state because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale. We have shown in section \ref{profiles} that atmosphere growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and so the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude. We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices. As in section \ref{analytic}, we define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass. At $t_{\rm{co}}$, runaway accretion commences.

In this section we explore the dependence of the crossover time on core, disk and gas parameters, and determine the minimum core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk. In section \ref{Mct} we show how the crossover time varies with core mass for fixed disk parameters, demonstrating that higher mass cores achieve runaway accretion more quickly. We explore how the crossover time is affected by by the mean molecular weight of the nebular gas in section \ref{muopacity} and by the disk temperature and pressure in section \ref{TPeffects}. Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates in section \ref{critcore}; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.

@@

We begin by generating atmosphere evolution profiles for cores of different masses under a fixed set of disk conditions. Figure \ref{fig:tvsM} displays $t_{\rm{co}}$ for $M_{\rm c}$ between 5 and 14 $M_{\oplus}$ at $a=10$ AU in our fiducial disk. The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity. This can also be seen from the analytical results in equation (\ref{eq:tcoolan}). As such, there is a minimum core mass required to form a planet within a fixed crossover time, e.g. the typical lifetime of the protoplanetary disk, which we defined above as the critical core mass. We defer numerical estimates of the critical core mass to section \ref{critcore}.

@@

 In this work we build evolutionary atmosphere profiles assuming an ideal gas equation of state with constant mean molecular weight and adiabatic index. A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization. Future work will include realistic equation of state tables, and we expect quantitatively different results. In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time. In Figure \ref{fig:tvsM} we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ from Figure \ref{fig:tvsM} with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen). For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}. However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long. Here we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model. We therefore estimate the analytic crossover time as the time needed for the atmosphere to reach an effective crossover mass $M_{\rm{atm}}=f M_{\rm c}$, where $f = 0.13$. The scaled analytic results are in good agreement with the numerical models. A lower mean molecular weight translates into a longer crossover time for the same core mass, since an atmosphere composed of heavier molecules has a smaller scale height, and therefore a weaker pressure gradient to oppose self-gravity.

@@

\subsubsection{Disk Temperature and Pressure}
\label{TPeffects}

@@

We find that the crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.

An increase in disk temperature results in a longer crossover time, as a higher temperature causes an increase in opacity. As a result, photons diffuse through the radiative layer more slowly, producing a lower planetary luminosity. Our analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation \ref{eq:tcoolan}), where $\beta$ is the constant in the opacity law (\ref{eq:opacitylaw}). For our choice of $\beta=2$, the crossover time is therefore strongly dependent on the opacity. Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.  We note, however, that the power law dependence of the crossover time on disk temperature found in our analytic model breaks down for the numerical model for low $T\di$. Nevertheless, the trends predicted by the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects} are still informative. For the full numerical solution, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$. The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.

From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to disk pressure. A higher disk pressure therefore results in a shorter crossover time. Increasing the disk pressure by a factor of $\sim 100$ only reduces the crossover time by a factor of $\sim 2$. As such, the crossover time is only weakly dependent on the nebular pressure.

@@

We can now combine the results obtained in section \ref{Mct} to determine the minimum core mass necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime, i.e. to have a crossover time less than the disk lifetime. We  denote this mass the \textit{critical core mass}.

%In the previous subsections we explored the effect of disk and gas properties on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass. The main question we address here is under what conditions the atmosphere can grow before the gas in the protoplanetary disk is dissipated. Specifically, we determine the minimum core mass that is necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime, i.e. to have a crossover time equal to the disk life timescale. We  denoted this mass the \textit{critical core mass} in section \ref{Mct}.

Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}). We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of disk temperature, pressure and mean molecular weight of the nebular gas.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
%\vspace{-0.5in}
\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for different mean molecular weight of the nebular gas. The analytic scaled results are plotted for comparison. The critical core mass is lower at larger stellocentric distances.}
\label{fig:Mcvsa}
\end{figure}

 In Figure \ref{fig:Mcvsa}, we see that the critical core mass decreases for larger separations, primarily due to the fact the disk temperature is lower in the outer regions of the disk. As we showed in section \ref{Mct}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass. The critical core mass is larger for a lower mean molecular weight, consistent with the results derived in section \ref{Mct}. We also explore the opacity effect on the critical core mass in Figure \ref{fig:Mcritopacity}. For an opacity, and therefore crossover time, that is an order of magnitude larger, the critical core mass is increased by roughly a factor of 2.

>>>>

%Standard calculations that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume a steady state at all times and find that, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass (see section \ref{intro}). In our model the atmosphere is no longer in steady state because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale. We have shown in section \ref{profiles} that atmosphere growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and so the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude. We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices. As in section \ref{analytic}, we define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass. At $t_{\rm{co}}$, runaway accretion commences.

In section \S\ref{coolingan} we defined the crossover time $t_{\rm co}$ as the time elapsed for the atmosphere to reach the crossover mass at $M_{\rm atm}=M\co$. Here we explore the dependence of the crossover time on core, disk and gas parameters, and determine the critical core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk. In section \ref{Mct} we show how the crossover time varies with core mass for fixed disk parameters, demonstrating that higher mass cores achieve runaway accretion more quickly. We explore how the crossover time is affected by by the mean molecular weight of the nebular gas in section \ref{muopacity} and by the disk temperature and pressure in section \ref{TPeffects}. Finally, we determine the critical core mass in section \ref{critcore} and explore its dependence on the dust opacity.

@@

We generate atmosphere evolution profiles for cores of different masses under a fixed set of disk conditions. Figure \ref{fig:tvsM} displays $t_{\rm{co}}$ for $M_{\rm c}$ between 5 and 14 $M_{\oplus}$ at $a=10$ AU in our fiducial disk. The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity. This is consistent with our analytic results in equation (\ref{eq:tcoolan}). As such, there is a minimum core mass required to form a planet within a fixed crossover time, e.g. the typical lifetime of the protoplanetary disk, which we defined as the critical core mass. We defer numerical estimates of the critical core mass to section \ref{critcore}.

@@

 In this work we build evolutionary atmosphere profiles assuming an ideal gas equation of state with constant mean molecular weight and adiabatic index. A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization. Future work will include realistic EOS tables, and we expect quantitatively different results. In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time. In Figure \ref{fig:tvsM} we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ from Figure \ref{fig:tvsM} with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen). For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}. However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long. Here we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model. We therefore estimate the analytic crossover time as the time needed for the atmosphere to reach an effective crossover mass $M_{\rm{atm}}=f M_{\rm c}$, where $f = 0.13$. The scaled analytic results are in good agreement with the numerical models. A lower mean molecular weight translates into a longer crossover time for the same core mass, since an atmosphere composed of heavier molecules has a smaller scale height, and therefore a weaker pressure gradient to oppose self-gravity.

@@

\subsubsection{Role of Disk Temperature and Pressure}
\label{TPeffects}

@@

The crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.

An increase in disk temperature results in a longer crossover time, as a higher temperature increases the dust opacity. As a result, photons diffuse through the radiative layer more slowly, producing a lower planetary luminosity. Our analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation \ref{eq:tcoolan}), where $\beta$ is the constant in the opacity law (\ref{eq:opacitylaw}). For our choice of $\beta=2$, the crossover time is therefore strongly dependent on the opacity. Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.  We note that the power law dependence of the crossover time on disk temperature found in our analytic model breaks down in the numerical model for low $T\di$. Nevertheless, the trends predicted by the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects} are still informative. For the full numerical solution, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$. The atmosphere evolution is thus highly dependent on the temperature in the protoplanetary disk.

From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to disk pressure, and hence a higher disk pressure results in a shorter crossover time. Increasing the disk pressure by a factor of $\sim 100$ only reduces the crossover time by a factor of $\sim 2$.  The crossover time is therefore only weakly dependent on the nebular pressure.

@@

We can now combine the results obtained in section \ref{Mct} to determine the minimum core mass to form a giant planet before the dissipation of the disk,. i.e. the critical core mass. Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}). We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of disk temperature, pressure and mean molecular weight of the nebular gas.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
%\vspace{-0.5in}
\caption{The critical core mass for a disk lifetime $\tau \sim 3$ Myrs as a function of semi-major axis, for different mean molecular weight of the nebular gas. The scaled analytic results are plotted for comparison. The critical core mass is lower at larger stellocentric distances.}
\label{fig:Mcvsa}
\end{figure}

Figure \ref{fig:Mcvsa}  shows that the critical core mass decreases for larger separations, primarily because the disk temperature is lower in the outer disk. As shown in section \ref{Mct}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass. The critical core mass is larger for a lower mean molecular weight, consistent with the results derived in section \ref{Mct}.  Figure \ref{fig:Mcritopacity} shows the effect of opacity reduction on the critical core mass. For an opacity one order of magnitude lower than the standard ISM opacity, the critical core mass decreases by roughly a factor of two.

### 2013-10-23 tex/draft/draft4.tex
draft4 tex + pdf, but several other files added automatically when trying to merge the pdf

The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is helpful for understanding atmospheric growth.  The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.  This insensitivity to the upper mass threshold is characteristic of accelerating growth.  We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.

@@

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.  Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate. In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}. The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.

>>>>

The fact that growth times peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth. In contrast with our analytic model, where the time to runaway accretion cannot be calculated self-consistently, the numerical model provides information about when unstable collapse occurs, i.e. when growth times decrease. It follows that the time to runaway accretion, $t_{\rm run}$, can be estimated as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to a fraction $f$ of its maximum value. This allows for a more rigorous estimate of the runaway accretion timescale, and thus of the critical core mass, as opposed to choosing $M_{\rm atm} \sim M\co$ as the criterion for collapse. In practice, however, the continuous accelerated growth in the late stages of the atmosphere evolution renders the estimated timescale to envelope collapse, and therefore the critical core mass, insensitive to the exact choice of runaway accretion time or atmosphere mass.  Since $M_{\rm atm} < M\co$ when growth times peak, we may proceed by analogy with the analytic model and set $t=t_{\rm co}$, when $M_{\rm atm}=M\co$, as the criterion for collapse. We note that in some cases our cooling model breaks down before the crossover mass, as explained in section \S\ref{sec:endoftime}, hence evolution is stopped before $M_{\rm atm}=M\co$. However, this breakdown occurs for $M_{\rm atm} \sim 0.8-0.9 M\co$, comparable to the crossover mass, so the additional time to reach crossover is negligible. It follows that $t_{\rm run} \approx t_{\rm co}$. By analogy with the analytic results we thus choose $t_{\rm co}$ as the timescale to runaway accretion, as it will give the same numerical results as $t_{\rm run}$ and it is easier to understand in the context of other core accretion models.

%The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.  This insensitivity to the upper mass threshold is characteristic of accelerating growth.

%We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.

@@

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.  Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate. As the breakdown always occurs after growth times peak, it does not affect our estimate of the timescale to runaway accretion $t_{\rm run}$. In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}. The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.
